\documentclass[12pt]{amsart}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue,linkcolor=blue]{hyperref}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Operators
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\MSE}{MSE}
\DeclareMathOperator{\relu}{ReLU}
\DeclareMathOperator{\spn}{span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Fisher}{Fisher}
\DeclareMathOperator{\softmax}{softmax}

% Shortcuts
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bJ}{\mathbf{J}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bF}{\mathbf{F}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bP}{\mathbf{P}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bxi}{\boldsymbol{\xi}}
\newcommand{\bmu}{\boldsymbol{\mu}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\abs}[1]{\left\lvert #1 \right\rvert}
\newcommand{\ip}[2]{\langle #1, #2 \rangle}

\title[Cross-Entropy Noise Propagation and Neural Collapse]{%
  A Cross-Entropy Noise Propagation Operator and the\\
  Effective Rank of the Class-Averaged Block at Neural Collapse}

\author{Lightman Chang}
\address{Independent Researcher}
\email{lightman.chang@gmail.com}

\date{\today}

\subjclass[2020]{Primary 68T07; Secondary 62G05, 41A25, 60J60, 62H30}

\keywords{Cross-entropy loss, Neural Collapse, Fisher information,
  noise propagation, benign overfitting, effective rank,
  multi-class classification}

\begin{document}

\begin{abstract}
We extend the noise-propagation framework
of~\cite{paperA}---originally formulated for the squared loss in
regression---to multi-class classification under the cross-entropy
loss. The technical step is to derive a Fisher-corrected noise
propagation operator $\bM_{\mathrm{CE}}$ directly from the
cross-entropy Hessian
$\nabla_{\theta}^{2}\ell_{\mathrm{CE}} = \bJ^{\top}\bH\bJ$, where
$\bH(\bx) = \diag(\bp(\bx)) - \bp(\bx)\bp(\bx)^{\top}$ is the
output-space Fisher matrix at the softmax probabilities $\bp(\bx)$.
Our first result (Theorem~\ref{thm:ce-trichotomy}) is that
$\tr(\bM_{\mathrm{CE}}) < \infty$ if and only if the underlying
power-law exponents satisfy $\beta > \alpha + 1/2$, identical to the
trichotomy for MSE. We then study the limit of $\bM_{\mathrm{CE}}$
along a sequence of trained predictors approaching the Neural
Collapse fixed point of Papyan--Han--Donoho. Although
$\bH$ becomes singular in the strict one-hot limit, the limiting
operator is well-defined on the $(K-1)$-dimensional zero-sum
subspace fixed by the simplex equiangular tight frame.
Theorem~\ref{thm:ce-rank} establishes two facts about the
effective rank on this subspace: (a) for the
\emph{class-averaged} block $\overline{\bM}_{\mathrm{CE}}$, the
$S_{K}$-symmetry of the simplex ETF combined with the irreducibility
of $\mathcal{Z}_{K}$ as an $S_{K}$-representation forces a scalar
operator, so $r_{\mathrm{eff}}(\overline{\bM}_{\mathrm{CE}})$ equals
exactly $K-1$; (b) for an individual per-sample block
$\bM_{\mathrm{CE}}^{(i)}$, the stabilizer is only $S_{K-1}$ and
$\mathcal{Z}_{K}$ splits into a $1$-dimensional ``target vs. mean
non-target'' direction and a $(K-2)$-dimensional ``differences among
non-targets'' subspace, on which $\bH$ has two distinct eigenvalues
$Ka b$ and $b$; consequently $r_{\mathrm{eff}}(\bM_{\mathrm{CE}}^{(i)})
\in [1, K-1]$ in general, with $r_{\mathrm{eff}} \to K-1$ only in
the symmetric edge case where the two eigenvalue blocks coincide.
The class-averaged statement gives a dimension-free generalization
scaling for $K$-class classification at the terminal phase of
training.
\end{abstract}

\maketitle

%% ============================================================
\section{Introduction}\label{sec:intro}
%% ============================================================

The unified bound of~\cite{paperA} on the test risk of
overparameterized regression involves a noise-propagation operator
$\bM = \bSigma^{-1}\bGamma\bSigma^{-1}$ built from the SVD of the
training Jacobian $\bJ$ and the test--train cross-correlation
$\bGamma$ of gradient features. The trichotomy
$\tr(\bM) < \infty \iff \beta > \alpha + 1/2$, under power-law
decay $\sigma_{j} \asymp j^{-\alpha}$ and
$\varrho_{j} := \sqrt{\bGamma_{jj}} \asymp j^{-\beta}$, characterizes
benign, tempered, and catastrophic regimes for squared loss; the
power-law model is standard in the random-feature analysis
of~\cite{mei2022}, with bias--variance refinements
in the companion~\cite{paperB}.

The same generalization story should apply to classification, but
the squared-loss derivation does not directly transfer: cross-entropy
introduces a nonlinear softmax link, and its parameter Hessian is
$\bJ^{\top}\bH\bJ$ rather than $\bJ^{\top}\bJ$. The natural question
is whether a Fisher-corrected operator preserves the trichotomy and
admits an exact effective-rank calculation at the Neural Collapse
(NC) fixed point of~\cite{papyan2020}.

This paper answers both questions affirmatively. We define
$\bM_{\mathrm{CE}}$ as the natural generalization of $\bM$ obtained
by replacing $\bJ^{\top}\bJ$ with the cross-entropy Hessian (see
Definition~\ref{def:M-CE} and the derivation that precedes it).
We then show:

\begin{enumerate}[label=(\roman*)]
\item The CE trichotomy
$\tr(\bM_{\mathrm{CE}}) < \infty \iff \beta > \alpha + 1/2$
holds whenever the predicted probabilities are bounded away from the
one-hot vertices uniformly along the relevant directions
(Theorem~\ref{thm:ce-trichotomy}).

\item Approaching the NC simplex equiangular tight frame fixed
point, the \emph{class-averaged} block
$\overline{\bM}_{\mathrm{CE}}$ on the zero-sum subspace
$\mathcal{Z}_{K}$ has effective rank exactly $K-1$
(Theorem~\ref{thm:ce-rank}\,(b)). The proof uses Schur's lemma:
$\mathcal{Z}_{K}$ is the $(K-1)$-dimensional standard irreducible
$S_{K}$-representation, and class-averaging produces an
$S_{K}$-equivariant operator, hence a scalar.
\item For a fixed \emph{per-sample} block $\bM_{\mathrm{CE}}^{(i)}$
the stabilizer of class $c_{i}$ is only $S_{K-1}$, which splits
$\mathcal{Z}_{K}$ into a $1$-dimensional and a $(K-2)$-dimensional
isotypic component carrying generically distinct eigenvalues, so
$r_{\mathrm{eff}}(\bM_{\mathrm{CE}}^{(i)}) \in [1, K-1]$ in
general; we give the explicit two-eigenvalue formula
(Theorem~\ref{thm:ce-rank}\,(a)).
\end{enumerate}

The first result extends the trichotomy across loss functions. The
second and third together identify the right object whose effective
rank is dimension-free at NC: the class-averaged block, not the
per-sample block. The averaged statement is what controls
generalization at the terminal phase, since the noise-leakage
variance $\sigma^{2}\tr(\bM_{\mathrm{CE}})$ in
Lemma~\ref{lem:noise-leakage}, and the corresponding effective rank,
aggregate over training samples; class-balance averages the
per-sample contributions back into a fully $S_{K}$-symmetric object.
We work conditional on the presence of NC and characterize what
this implies for the effective rank, rather than proving NC.

\paragraph{Relation to prior work.}
The Neural Collapse phenomenon was identified empirically
in~\cite{papyan2020} and characterized mathematically in
the unconstrained-features setting under cross-entropy
by~\cite{lu2022, mixon2022} and under squared loss / central path
by~\cite{han2022}. The class-imbalanced extension of unconstrained
features is~\cite{Fang2021}. The cross-entropy Fisher matrix
$\bH = \diag(\bp) - \bp\bp^{\top}$ and its degeneracy at one-hot
vertices is classical (see~\cite{Amari2016}); we use only the
finite-rank facts about $\bH$ and the simplex ETF geometry.

\paragraph{Organization.}
Section~\ref{sec:setting} fixes notation and recalls the
classification setting. Section~\ref{sec:operator} derives
$\bM_{\mathrm{CE}}$ from the CE Hessian. Section~\ref{sec:trichotomy}
proves the CE trichotomy. Section~\ref{sec:nc} derives the
effective-rank formula at the NC fixed point. Section~\ref{sec:disc}
discusses limitations.

%% ============================================================
\section{Setting and notation}\label{sec:setting}
%% ============================================================

Let $K \geq 2$ be the number of classes. The predictor has
$K$-dimensional output
$f(\bx;\theta) = (f_{1}(\bx;\theta),\ldots,f_{K}(\bx;\theta)) \in \R^{K}$,
softmax probabilities
$p_{k}(\bx;\theta) = e^{f_{k}}/\sum_{k'} e^{f_{k'}}$, and one-hot
labels $\by_{i} \in \{e_{1},\ldots,e_{K}\}$, where $e_{k}$ is the
$k$-th standard basis vector. Training data
$S = \{(\bx_{i},\by_{i})\}_{i=1}^{n}$ are i.i.d. The cross-entropy
loss is
\[
  \ell_{\mathrm{CE}}(f,\by) = -\sum_{k=1}^{K} y_{k}\log p_{k}(f).
\]
A direct computation gives the well-known output-space Hessian
\begin{equation}\label{eq:Hbox}
  \nabla_{f}^{2}\ell_{\mathrm{CE}}(f,\by) = \bH(\bx) := \diag(\bp(\bx)) - \bp(\bx)\bp(\bx)^{\top},
\end{equation}
which is symmetric positive semidefinite, has rank $K-1$ for any
$\bp$ in the open simplex (it has a single zero eigenvalue along
$\mathbf{1}$), and rank $0$ at every one-hot vertex
$\bp = e_{k}$ (since at $\bp = e_{k}$ we have
$\diag(e_{k}) = e_{k}e_{k}^{\top}$, hence
$\bH = \diag(e_{k}) - e_{k}e_{k}^{\top} = 0$).
The Fisher matrix $\bH(\bx)$ is the building block of the parameter
Hessian:
\begin{equation}\label{eq:CE-Hessian}
  \nabla_{\theta}^{2}\ell_{\mathrm{CE}}(\bx,\by;\theta)
  = \bJ(\bx)^{\top}\bH(\bx)\bJ(\bx) + \text{(label-dependent rank-1 term)},
\end{equation}
where $\bJ(\bx) \in \R^{K\times p}$ is the per-sample Jacobian of
$f(\bx;\cdot)$ with respect to $\theta$. The label-dependent term
vanishes in expectation over labels at any predictor that is exact
on the population (or at any limit point with vanishing gradient),
and we drop it from this point onward.

We adopt the notation of~\cite{paperA}. Let
$\bJ_{\mathrm{CE}} \in \R^{nK \times p}$ denote the stacked
training Jacobian with rows
$\nabla_{\theta} f_{k}(\bx_{i};\theta^{*})^{\top}$ ranging over
$(i,k) \in [n]\times[K]$, and let
$\bH = \bigoplus_{i=1}^{n}\bH(\bx_{i})$ denote the block-diagonal
sample Fisher. Denote the cross-class test--train cross-correlation
\begin{equation}\label{eq:GammaCE}
  (\bGamma_{\mathrm{CE}})_{(j,k),(j',k')} = \E_{\bx}[\psi_{j,k}(\bx)\psi_{j',k'}(\bx)],
  \quad \psi_{j,k}(\bx) = \nabla_{\theta} f_{k}(\bx;\theta^{*})^{\top}\bv_{j},
\end{equation}
where $\{\bv_{j}\}$ is the right SVD basis of an appropriate factor
of $\bJ_{\mathrm{CE}}$ (see Section~\ref{sec:operator}).

%% ============================================================
\section{Deriving $\bM_{\mathrm{CE}}$ from the cross-entropy
  Hessian}\label{sec:operator}
%% ============================================================

In MSE, the operator $\bM = \bSigma^{-1}\bGamma\bSigma^{-1}$ arises
because the test risk decomposes as a quadratic form in the
predicted residuals, weighted by the inverse of $\bJ^{\top}\bJ$ at
the interpolant. Concretely, $\bSigma$ comes from the SVD of the
training Jacobian, and the noise-leakage variance equals
$\sigma^{2}\tr(\bSigma^{-1}\bGamma\bSigma^{-1})$.

For cross-entropy at a (near-)interpolating predictor, the
linearized test risk is governed by the inverse of the parameter
Hessian~\eqref{eq:CE-Hessian}, that is
$(\bJ_{\mathrm{CE}}^{\top}\bH\bJ_{\mathrm{CE}})^{-1}$. Two natural
factorizations arise:

\begin{enumerate}
\item Define the \emph{Fisher-weighted Jacobian}
\[
  \widetilde\bJ := \bH^{1/2}\bJ_{\mathrm{CE}} \in \R^{nK\times p},
\]
with SVD $\widetilde\bJ = \widetilde\bU\,\widetilde\bSigma\,\widetilde\bV^{\top}$.
Then
$\bJ_{\mathrm{CE}}^{\top}\bH\bJ_{\mathrm{CE}} = \widetilde\bV\,\widetilde\bSigma^{2}\,\widetilde\bV^{\top}$,
so the parameter Hessian's eigenstructure coincides with that of
$\widetilde\bJ^{\top}\widetilde\bJ$.

\item The right singular vectors $\widetilde\bV$ provide the natural
basis for gradient features
$\widetilde\psi_{j,k}(\bx) := \nabla_{\theta}f_{k}(\bx;\theta^{*})^{\top}\widetilde\bv_{j}$,
and the corresponding cross-correlation matrix is
$\widetilde\bGamma_{(j,k),(j',k')} = \E_{\bx}[\widetilde\psi_{j,k}(\bx)\widetilde\psi_{j',k'}(\bx)]$.
The matrix $\widetilde\bSigma_{\mathrm{CE}}$ refers throughout to
the singular value matrix $\widetilde\bSigma$ above; these are the
square roots of the eigenvalues of the parameter Hessian.
\end{enumerate}

These two ingredients give the operator that propagates label noise
through the linearized CE risk:

\begin{definition}[Cross-entropy noise propagation operator]\label{def:M-CE}
The \emph{cross-entropy noise propagation operator} is
\begin{equation}\label{eq:M-CE}
  \bM_{\mathrm{CE}}
  := \widetilde\bSigma^{-1}\,\widetilde\bGamma\,\widetilde\bSigma^{-1},
  \qquad \widetilde\bJ = \bH^{1/2}\bJ_{\mathrm{CE}} = \widetilde\bU\widetilde\bSigma\widetilde\bV^{\top},
\end{equation}
where the inverse on the right is the Moore--Penrose pseudoinverse
restricted to the range of $\widetilde\bSigma$.
\end{definition}

\begin{definition}[Effective rank]\label{def:reff}
For any non-zero positive semidefinite operator $\bA$ on a finite-dimensional
inner-product space, the \emph{effective rank} is
\[
  r_{\mathrm{eff}}(\bA) \;:=\; \frac{(\tr\bA)^{2}}{\norm{\bA}_{F}^{2}}.
\]
This quantity satisfies $1 \leq r_{\mathrm{eff}}(\bA) \leq \rank(\bA)$,
with equality on the right iff all nonzero eigenvalues are equal,
and is invariant under positive scalar rescaling of $\bA$.
\end{definition}

\begin{remark}[Equivalence with the form
$\bF^{-1}\bGamma\bF^{-1}$]\label{rem:M-CE-eq}
Some prior writeups state $\bM_{\mathrm{CE}} = \bF^{-1}\bGamma_{\mathrm{CE}}\bF^{-1}$
with $\bF = \bH^{1/2}\bSigma_{\mathrm{CE}}$, where $\bSigma_{\mathrm{CE}}$
denotes the singular values of $\bJ_{\mathrm{CE}}$. This is the same
operator as~\eqref{eq:M-CE} written in the basis of $\bJ_{\mathrm{CE}}$
rather than $\widetilde\bJ$: $\widetilde\bSigma$ are precisely the
singular values of $\bH^{1/2}\bJ_{\mathrm{CE}}$, which equal the
square roots of the eigenvalues of
$\bJ_{\mathrm{CE}}^{\top}\bH\bJ_{\mathrm{CE}}$. We retain
Definition~\ref{def:M-CE} as the canonical form because the
factorization is unambiguous.
\end{remark}

\paragraph{Cross-reference to \cite{paperC}.}
The symbols $\bM, \bSigma, \bGamma$ used in this paper are the
direct cross-entropy analogs of the squared-loss operators in the
companion paper~\cite{paperC}. The dictionary is:
$\bSigma$ (singular values of $\bJ$) $\leftrightarrow$
$\widetilde\bSigma$ (singular values of
$\widetilde\bJ = \bH^{1/2}\bJ_{\mathrm{CE}}$);
$\bGamma$ (test--train cross-correlation in the right singular
basis of $\bJ$) $\leftrightarrow$ $\widetilde\bGamma$ (in the right
singular basis of $\widetilde\bJ$); and
$\bM = \bSigma^{-1}\bGamma\bSigma^{-1}$ $\leftrightarrow$
$\bM_{\mathrm{CE}} = \widetilde\bSigma^{-1}\widetilde\bGamma\widetilde\bSigma^{-1}$.
Setting $\bH \equiv \bI$ in the cross-entropy definitions recovers
the squared-loss objects of~\cite{paperC} verbatim.

The justification for Definition~\ref{def:M-CE} is the following
linearization lemma, which derives the noise-leakage variance from
a Taylor expansion of the empirical CE risk near $\theta^{*}$. The
argument is the cross-entropy analog of the MSE template
in~\cite[\S 2]{paperA}; the loss-specific ingredient is the
substitution of the CE Hessian $\bJ^{\top}\bH\bJ$ for the squared-loss
Hessian $\bJ^{\top}\bJ$, equivalently the replacement of $\bJ$ by
$\widetilde\bJ = \bH^{1/2}\bJ$.

\begin{lemma}[Linearized noise leakage under CE]\label{lem:noise-leakage}
Let $\theta^{*}$ be a stationary point of the empirical CE risk on
the training set $S$, with parameter Hessian
$\bF := \bJ_{\mathrm{CE}}^{\top}\bH\bJ_{\mathrm{CE}}$ invertible
on its range. Suppose label perturbations
$\bxi_{i} \in \R^{K}$ satisfy $\E[\bxi_{i}] = 0$ and
$\E[\bxi_{i}\bxi_{i}^{\top}] = \sigma^{2}\,\bH(\bx_{i})$ (the
Fisher-weighted noise model that matches the CE link). Let
$\theta^{*}_{\bxi}$ denote the perturbed stationary point. Then to
linear order in $\bxi$,
\begin{enumerate}[label=(\roman*)]
\item the parameter perturbation is
$\delta\theta = \bF^{\dagger}\bJ_{\mathrm{CE}}^{\top}\bH^{1/2}\bxi
              = \widetilde\bV\widetilde\bSigma^{-1}\widetilde\bU^{\top}\bxi$;
\item along any gradient feature direction $\widetilde\bv_{j}$
the variance contribution to the test residual at $\bx_{*}$ equals
$\sigma^{2}\widetilde\bGamma_{jj}/\widetilde\sigma_{j}^{2}$;
\item summing,
$\Var(\text{linearized test residual}) = \sigma^{2}\tr(\bM_{\mathrm{CE}})$.
\end{enumerate}
\end{lemma}

\begin{proof}
Linearize the empirical-risk gradient near $\theta^{*}$:
\[
  \nabla_{\theta}\widehat{\mathcal{R}}_{\mathrm{CE}}(\theta^{*} + \delta\theta;\, \by + \bxi)
   = \bF\,\delta\theta - \bJ_{\mathrm{CE}}^{\top}\bH^{1/2}\bxi
   + O(\norm{\delta\theta}^{2} + \norm{\bxi}^{2}),
\]
using $\nabla_{\theta}\ell_{\mathrm{CE}}(\bx,\by;\theta) = \bJ(\bx)^{\top}(\bp(\bx;\theta) - \by)$
and the noise model. Setting the gradient to zero gives
$\delta\theta = \bF^{\dagger}\bJ_{\mathrm{CE}}^{\top}\bH^{1/2}\bxi$,
which is~(i); substituting the SVD
$\widetilde\bJ = \widetilde\bU\widetilde\bSigma\widetilde\bV^{\top}$
into $\bF = \widetilde\bV\widetilde\bSigma^{2}\widetilde\bV^{\top}$
yields the explicit form. For (ii), the test residual contribution
along $\widetilde\bv_{j}$ is
$\widetilde\psi_{j,\cdot}(\bx_{*})^{\top}\delta\theta$, whose
variance is
$\sigma^{2}\widetilde\sigma_{j}^{-2}\widetilde\bGamma_{jj}$ by
orthonormality of $\widetilde\bU$. For (iii), sum to recognize
$\sum_{j}\widetilde\bGamma_{jj}/\widetilde\sigma_{j}^{2}
   = \tr(\widetilde\bSigma^{-1}\widetilde\bGamma\widetilde\bSigma^{-1})
   = \tr(\bM_{\mathrm{CE}})$.
\end{proof}

This lemma plays the same role as the squared-loss derivation
in~\cite{paperA}: $\bM_{\mathrm{CE}}$ is exactly the quadratic form
that controls noise-leakage variance under CE, with $\bH$ replacing
the identity in the output-space inner product.

%% ============================================================
\section{Cross-entropy trichotomy}\label{sec:trichotomy}
%% ============================================================

\begin{theorem}[Cross-entropy trichotomy]\label{thm:ce-trichotomy}
Suppose the singular values $\widetilde\sigma_{j}$ of $\widetilde\bJ
= \bH^{1/2}\bJ_{\mathrm{CE}}$ and the diagonal entries
$\widetilde\bGamma_{jj}$ (in the right singular basis of
$\widetilde\bJ$) exhibit power-law decay
$\widetilde\sigma_{j} \asymp j^{-\alpha}$,
$\widetilde\varrho_{j} := \sqrt{\widetilde\bGamma_{jj}} \asymp j^{-\beta}$.
Suppose furthermore that the Fisher block is uniformly non-degenerate
in the sense of the operator inequality
\begin{equation}\label{eq:hmin-op}
  \bH(\bx_{i}) \;\succeq\; h_{\min}\,\bP_{\mathcal{Z}_{K}}
  \qquad \text{for all } i = 1,\ldots,n,
\end{equation}
for some $h_{\min} > 0$, where $\bP_{\mathcal{Z}_{K}}$ is the
orthogonal projection onto the zero-sum subspace
$\mathcal{Z}_{K} = \{u \in \R^{K} : \mathbf{1}^{\top}u = 0\}$.
Then
\[
  \tr(\bM_{\mathrm{CE}}) < \infty \iff \beta > \alpha + \tfrac{1}{2}.
\]
\end{theorem}

\begin{proof}
By Definition~\ref{def:M-CE},
\[
  \tr(\bM_{\mathrm{CE}})
  = \sum_{j}\widetilde\bGamma_{jj}/\widetilde\sigma_{j}^{2}
  \asymp \sum_{j} j^{-2\beta}/j^{-2\alpha}
  = \sum_{j} j^{2(\alpha-\beta)},
\]
which converges iff $2(\alpha-\beta) < -1$, i.e.\ $\beta > \alpha + 1/2$.
This is a statement about the singular values of
$\widetilde\bJ$ and the diagonal of $\widetilde\bGamma$ in
$\widetilde\bJ$'s right singular basis, so no basis conversion is
needed. Hypothesis~\eqref{eq:hmin-op} ensures that
$\widetilde\bJ = \bH^{1/2}\bJ_{\mathrm{CE}}$ has full rank on the
zero-sum subspace whenever $\bJ_{\mathrm{CE}}$ does, with smallest
nonzero singular value at least $h_{\min}^{1/2}$ times that of
$\bJ_{\mathrm{CE}}$ on $\mathcal{Z}_{K}$, so the power-law
hypotheses are equivalent to power laws on $\bJ_{\mathrm{CE}}$ up to
a multiplicative constant.
\end{proof}

\begin{remark}[On the uniform Fisher lower bound]\label{rem:hmin}
Hypothesis~\eqref{eq:hmin-op} is the cross-entropy analog of the
``well-conditioned predictor'' assumption: at any predictor whose
softmax outputs $\bp(\bx)$ avoid the boundary of the simplex, the
nonzero eigenvalues of $\bH(\bx)$ are bounded below by a positive
constant on $\mathcal{Z}_{K}$. For a stable training trajectory
away from the zero-margin classifier, this is automatic with
$h_{\min}$ controlled by the entropy of the outputs. The hypothesis
fails in the strict NC limit; we treat that limit separately in
Section~\ref{sec:nc} via a different argument.
\end{remark}

\begin{remark}\label{rem:Sigma-CE}
The matrix $\bSigma_{\mathrm{CE}}$ in the form
$\bM_{\mathrm{CE}} = \bF^{-1}\bGamma_{\mathrm{CE}}\bF^{-1}$ of
Remark~\ref{rem:M-CE-eq} denotes the diagonal matrix of singular
values of the (unweighted) Jacobian $\bJ_{\mathrm{CE}}$; under
hypothesis~\eqref{eq:hmin-op}, the singular values of
$\widetilde\bJ$ relate to those of $\bJ_{\mathrm{CE}}$ via
$\widetilde\sigma_{j} \in [h_{\min}^{1/2}\sigma_{j}^{\mathrm{CE}},\, h_{\max}^{1/2}\sigma_{j}^{\mathrm{CE}}]$
where $h_{\max} = \norm{\bH}_{\mathrm{op}}$.
\end{remark}

%% ============================================================
\section{Effective rank at the Neural Collapse fixed
  point}\label{sec:nc}
%% ============================================================

We now compute the effective rank
$r_{\mathrm{eff}}(\bM_{\mathrm{CE}})$ in the limit where the
trained predictor approaches the Neural Collapse fixed point
of~\cite{papyan2020}. A subtlety: in the strict infinite-margin
limit (logits diverging in the simplex ETF directions), the softmax
output for each sample becomes one-hot, $\bH(\bx) \to 0$, and
$\bM_{\mathrm{CE}}$ degenerates. We address this by considering a
sequence of predictors whose last-layer features approach the NC
fixed point with \emph{uniformly bounded logit norms}; in this
regime the limiting $\bp(\bx)$ is a probability vector strictly in
the open simplex, peaked on the correct class but with positive
mass on every coordinate. This is the regime relevant to NC
phenomenology at any finite training time, and the one in which the
unconstrained-features analyses of~\cite{lu2022,mixon2022,Fang2021}
yield well-defined limits. The strict-margin limit is treated
separately by rescaling (Remark~\ref{rem:strict-NC}).

\subsection{Geometry of the simplex equiangular tight frame}

Recall the NC fixed point: the last-layer features satisfy
$h(\bx_{i}) = \mu_{c(\bx_{i})}$ where $\{\mu_{c}\}_{c=1}^{K}$
forms a simplex equiangular tight frame in $\R^{K-1}$ with
$\sum_{c}\mu_{c} = 0$ and
$\mu_{c}^{\top}\mu_{c'} = \tfrac{K\delta_{cc'} - 1}{K-1}\,\norm{\mu}^{2}$
for some common norm $\norm{\mu}$. The classifier weights
$\{\bw_{c}\}$ likewise satisfy
$\bw_{c} = (\norm{\bw}/\norm{\mu})\mu_{c}$ up to a scalar.

The image of the per-class logit difference map
$h \mapsto (\bw_{1}^{\top}h,\ldots,\bw_{K}^{\top}h)$ at the NC fixed
point therefore lies in the $(K-1)$-dimensional zero-sum subspace
\[
  \mathcal{Z}_{K} := \{u \in \R^{K} : \mathbf{1}^{\top}u = 0\}.
\]
Let $\bP_{\mathcal{Z}}$ be the orthogonal projection onto
$\mathcal{Z}_{K}$, that is
$\bP_{\mathcal{Z}} = \bI_{K} - \tfrac{1}{K}\mathbf{1}\mathbf{1}^{\top}$.

\subsection{Limiting Fisher and limiting cross-correlation}

\begin{lemma}[Fisher matrix under bounded-logit NC]\label{lem:H-NC}
Consider a sequence of predictors $\{\theta^{(t)}\}$ whose
last-layer features approach the NC fixed point and whose logit
vectors $f(\bx_{i};\theta^{(t)}) \in \R^{K}$ satisfy
$\norm{f(\bx_{i};\theta^{(t)})}_{\infty} \leq L$ for some finite
$L > 0$. Then for any sample $\bx_{i}$ and along any limit point of
$\{\theta^{(t)}\}$, every coordinate of the predicted probability
$\bp(\bx_{i};\theta^{(t)})$ is bounded below by
$p_{\min} := e^{-2L}/K > 0$, the matrix $\bH(\bx_{i})$ has rank
exactly $K-1$ with kernel $\spn(\mathbf{1})$, and the operator
inequality
\[
  \bH(\bx_{i}) \;\succeq\; h_{\min}\,\bP_{\mathcal{Z}_{K}},
  \qquad h_{\min} := (K-1)\,p_{\min}^{2},
\]
holds uniformly in $i$ and $t$.
\end{lemma}

\begin{proof}
For any logit vector $f$ with $\norm{f}_{\infty} \leq L$, the
softmax probabilities satisfy $p_{k} = e^{f_{k}}/\sum_{k'}e^{f_{k'}} \geq e^{-L}/(Ke^{L}) = e^{-2L}/K$,
giving $p_{\min} > 0$. The Fisher matrix
$\bH = \diag(\bp) - \bp\bp^{\top}$ satisfies
$\bH\mathbf{1} = \bp - \bp(\mathbf{1}^{\top}\bp) = 0$, so
$\spn(\mathbf{1}) \subseteq \ker\bH$, with equality on the open
simplex (a standard fact about the multinomial Fisher matrix; see
e.g.~\cite[\S 3]{Amari2016}). For the lower eigenvalue bound on
$\mathcal{Z}_{K}$, the classical estimate is
\[
  \lambda_{\min}\bigl(\bH|_{\mathcal{Z}_{K}}\bigr)
  \;\geq\; p_{\min}\,(1 - p_{\max}),
\]
which follows from the variational characterization. Since
$p_{\max} = 1 - \sum_{k \neq \arg\max}p_{k} \leq 1 - (K-1)p_{\min}$,
we obtain
\[
  \lambda_{\min}\bigl(\bH|_{\mathcal{Z}_{K}}\bigr)
  \;\geq\; p_{\min}\,\bigl(1 - (1 - (K-1)p_{\min})\bigr)
  \;=\; (K-1)\,p_{\min}^{2}.
\]
(The chain ``$p_{\max} \leq 1 - (K-1)p_{\min} \leq 1 - 1/K$'' that
might be tempting requires $p_{\min} \geq 1/K$ and is generally
false: $p_{\min} = e^{-2L}/K < 1/K$ whenever $L > 0$. Hence we
retain the sharper bound $h_{\min} = (K-1)p_{\min}^{2}$, which is
correct for all $L > 0$.)
\end{proof}

\begin{remark}[Strict-one-hot limit and rescaling]\label{rem:strict-NC}
In the strict infinite-margin limit ($L \to \infty$), the
bounded-logit condition fails, $\bp \to e_{c}$, and $\bH \to 0$, so
$\widetilde\bJ \to 0$ and Definition~\ref{def:M-CE} degenerates as
written. The natural rescaled object is
\[
  \widetilde\bJ_{\mathrm{rsc}} := h_{\min}^{-1/2}\widetilde\bJ
   \;=\; (h_{\min}^{-1/2}\bH^{1/2})\,\bJ_{\mathrm{CE}}.
\]
The operator $\bM_{\mathrm{CE}}$ rebuilt from
$\widetilde\bJ_{\mathrm{rsc}}$ has the same effective rank as the
unrescaled $\bM_{\mathrm{CE}}$, since $r_{\mathrm{eff}}$ is
invariant under positive scalar rescaling of $\widetilde\bSigma$. We
caution that $h_{\min}^{-1/2}\bH^{1/2}$ does \emph{not} converge to
the projector $\bP_{\mathcal{Z}_{K}}$: at NC the per-sample
probabilities $\bp(\bx_{i})$ are of the form $(a, b, \ldots, b)$
with $a + (K-1)b = 1$ and $a > b$, and the eigenvalues of
$\bH(\bx_{i})|_{\mathcal{Z}_{K}}$ are $K a b$ on the
$1$-dimensional ``target vs. mean non-target'' direction and $b$ on
the $(K-2)$-dimensional ``differences among non-targets'' subspace
(direct calculation, see Step~2 of the proof of
Theorem~\ref{thm:ce-rank}). The ratio $K a b / b = K a$ remains
finite (and equals $K$ in the strict $a \to 1$ limit), so
$h_{\min}^{-1/2}\bH^{1/2}$ rescales the trace but not the
eigenvalue ratio: rescaling by $h_{\min}$ does not symmetrize the
spectrum. This is what forces us to track the two eigenvalue blocks
separately in Theorem~\ref{thm:ce-rank}, rather than collapsing
them by a strict-NC limit. Below we work with finite logit scale
throughout.
\end{remark}

\subsection{Effective rank formula}

Before stating the theorem, we make the structural NC hypotheses
explicit. The simplex-ETF geometry of~\cite{papyan2020} fixes the
last-layer features and classifier weights, but the per-sample
Jacobian also depends on the backbone (everything below the last
layer). The following assumption, standard in the
unconstrained-features literature
\cite{lu2022,mixon2022,Fang2021}, isolates what is needed.

\begin{assumption}[NC backbone non-degeneracy]\label{ass:nc-backbone}
At $\theta^{*}$, with last-layer feature map
$h(\bx;\theta^{*}) \in \R^{K-1}$ and last-layer classifier
$W \in \R^{K\times(K-1)}$ in simplex-ETF configuration:
\begin{enumerate}[label=(\alph*)]
\item (Class collapse / NC1) For every sample $\bx_{i}$ in class
$c_{i} \in \{1,\ldots,K\}$, $h(\bx_{i};\theta^{*}) = \mu_{c_{i}}$
where $\{\mu_{c}\}$ is the simplex ETF
($\sum_{c}\mu_{c} = 0$,
$\mu_{c}^{\top}\mu_{c'} = (K\delta_{cc'} - 1)/(K-1) \cdot \norm{\mu}^{2}$).
\item (Self-duality / NC2) $W = (\norm{W}/\norm{\mu})\,M$ where
$M^{\top} = [\mu_{1}\,\cdots\,\mu_{K}]$.
\item (Backbone full-rank in ETF directions) The backbone Jacobian
$\partial h(\bx_{i};\theta^{*})/\partial\theta \in \R^{(K-1) \times p}$
has full row rank $K-1$ for every $i$.
\end{enumerate}
\end{assumption}

Hypothesis (c) is the chain-rule input to the per-sample Jacobian
$\bJ(\bx_{i}) \in \R^{K\times p}$; it follows in the
unconstrained-features regime where $h$ is a free variable
\cite{mixon2022}, and is the operative assumption in feature-learning
analyses of NC. We do not derive it from NC1--NC4; we assume it.

We measure the effective rank on the per-sample block (equivalently,
on the output-averaged operator), which is the geometric object
fixed by simplex-ETF symmetry and is independent of $n$.

\begin{definition}[Per-sample / averaged $\bM_{\mathrm{CE}}$ block]\label{def:M-bar}
Let $\bM_{\mathrm{CE}}^{(i)} \in \R^{K\times K}$ denote the
per-sample restriction of $\bM_{\mathrm{CE}}$ to the output block
indexed by sample $i$, viewed as an operator on $\R^{K}$. Define
the \emph{averaged operator}
\[
  \overline{\bM}_{\mathrm{CE}}
  := \frac{1}{n}\sum_{i=1}^{n}\bM_{\mathrm{CE}}^{(i)}.
\]
\end{definition}

\begin{theorem}[Effective rank at NC]\label{thm:ce-rank}
Let $\theta^{*}$ be a limit point of the training trajectory of
Lemma~\ref{lem:H-NC} satisfying Assumption~\ref{ass:nc-backbone},
and assume the training set is class-balanced ($n/K$ samples per
class).
\begin{enumerate}[label=(\alph*)]
\item (\textbf{Per-sample block, two-eigenvalue formula.}) For each
sample $i$ in class $c_{i}$, the per-sample block
$\bM_{\mathrm{CE}}^{(i)}$ vanishes on $\spn(\mathbf{1})$ and on
$\mathcal{Z}_{K}$ has at most two distinct eigenvalues:
\[
  \lambda_{1}^{(i)} \;=\; \frac{\gamma_{1}^{(i)}}{K\,a\,b}
  \quad\text{(multiplicity $1$, on the target-vs-non-target axis)},
\]
\[
  \lambda_{2}^{(i)} \;=\; \frac{\gamma_{2}^{(i)}}{b}
  \quad\text{(multiplicity $K-2$, on differences among non-targets)},
\]
where $a, b > 0$ are the target / non-target softmax probabilities
($a + (K-1)b = 1$), and $\gamma_{1}^{(i)}, \gamma_{2}^{(i)} > 0$ are
the corresponding eigenvalues of the per-sample
$\widetilde\bGamma$-block under the $S_{K-1}$-isotypic decomposition.
Consequently
\[
  r_{\mathrm{eff}}\bigl(\bM_{\mathrm{CE}}^{(i)}\bigr)
  \;=\;
  \frac{\bigl(\lambda_{1}^{(i)} + (K-2)\lambda_{2}^{(i)}\bigr)^{2}}
       {(\lambda_{1}^{(i)})^{2} + (K-2)(\lambda_{2}^{(i)})^{2}}
  \;\in\; [1,\,K-1],
\]
with the upper bound $K-1$ attained iff $\lambda_{1}^{(i)} = \lambda_{2}^{(i)}$
and the lower bound approached as one eigenvalue dominates the
other. (For $K=2$ the second block is absent and
$r_{\mathrm{eff}} = 1$; for $K \geq 3$ the range is non-degenerate.)
\item (\textbf{Class-averaged block, exact $K-1$.}) The class-averaged
operator $\overline{\bM}_{\mathrm{CE}}$ on $\mathcal{Z}_{K}$ is a
scalar:
\[
  \overline{\bM}_{\mathrm{CE}}\bigl|_{\mathcal{Z}_{K}}
  \;=\; \kappa\,\bP_{\mathcal{Z}_{K}}, \qquad \kappa > 0,
\]
hence
\[
  r_{\mathrm{eff}}\bigl(\overline{\bM}_{\mathrm{CE}}\bigr)
  \;=\; K - 1.
\]
\end{enumerate}
\end{theorem}

\begin{proof}
\emph{Step 1: Per-sample block has rank exactly $K-1$ on
$\mathcal{Z}_{K}$.}
By Lemma~\ref{lem:H-NC}, $\bH(\bx_{i})$ is positive definite on
$\mathcal{Z}_{K}$ and zero on $\spn(\mathbf{1})$. By
Assumption~\ref{ass:nc-backbone}(c) the backbone Jacobian has rank
$K-1$, and the chain rule gives
\[
  \bJ(\bx_{i}) \;=\; W \cdot \frac{\partial h(\bx_{i};\theta^{*})}{\partial\theta}
                \;+\; \frac{\partial f}{\partial W}\Big|_{(\bx_{i})} \cdot
                  \frac{\partial W}{\partial\theta},
\]
where $\partial f/\partial W$ at the NC weight has rank $K-1$ on
$\mathcal{Z}_{K}$ (the simplex-ETF rows of $W$ span $\mathcal{Z}_{K}$
exactly, by Assumption~\ref{ass:nc-backbone}(b)). Hence
$\bJ(\bx_{i})\bP_{\mathcal{Z}_{K}}$ has row rank exactly $K-1$, and
the same holds for $\widetilde\bJ(\bx_{i}) = \bH(\bx_{i})^{1/2}\bJ(\bx_{i})$.

\emph{Step 2: Eigenstructure of $\bH(\bx_{i})$ on $\mathcal{Z}_{K}$.}
At NC, the predicted probability for sample $\bx_{i}$ in class
$c_{i}$ is permutation-symmetric across the $K-1$ non-target
classes (since the corresponding logits
$\bw_{c}^{\top}\mu_{c_{i}} = (\norm{\bw}/\norm{\mu})\mu_{c}^{\top}\mu_{c_{i}}$
take the common value $-(\norm{\bw}\norm{\mu})/(K-1)$ for all
$c \neq c_{i}$ by the simplex-ETF inner product). Thus
$\bp(\bx_{i})$ has the form $(\,\ldots,\,a,\,\ldots\,)$ with the
target coordinate $a$ at position $c_{i}$ and every non-target
coordinate equal to $b$, with $a + (K-1)b = 1$ and $a > b > 0$
(strict inequalities by the bounded-logit Lemma~\ref{lem:H-NC} and
the assumption that NC is non-trivial, i.e., not the uniform
prediction $a = b = 1/K$).

Decompose $\mathcal{Z}_{K}$ into the two $S_{K-1}$-isotypic
components fixed by permutations of the non-target indices:
\begin{itemize}
\item $V_{1}^{(i)} \;:=\; \spn\bigl((K-1)\,e_{c_{i}} - \sum_{c \neq c_{i}}e_{c}\bigr)$,
the $1$-dimensional ``target vs. mean of non-targets'' direction;
\item $V_{2}^{(i)} \;:=\; \{u \in \mathcal{Z}_{K} : u_{c_{i}} = 0\}$,
the $(K-2)$-dimensional ``differences among non-target coordinates''
subspace.
\end{itemize}
A direct computation (using $a + (K-1)b = 1$, equivalently
$1 - (a-b) = K b$ and $1 + (K-1)(a-b) = K a$) gives
\[
  \bH(\bx_{i})\,v \;=\; K a b\,v \quad \forall v \in V_{1}^{(i)},
  \qquad
  \bH(\bx_{i})\,v \;=\; b\,v \quad \forall v \in V_{2}^{(i)}.
\]
For $v \in V_{2}^{(i)}$ this is immediate from
$\bH v = \diag(\bp)v - \bp(\bp^{\top}v) = b v - 0$ since
$\bp^{\top}v = b\sum_{c\neq c_{i}}v_{c} = 0$ on $V_{2}^{(i)}$.
For $v_{1} = (K-1)e_{c_{i}} - \sum_{c \neq c_{i}}e_{c} \in V_{1}^{(i)}$,
$\diag(\bp)v_{1} = (K-1)a\,e_{c_{i}} - b\sum_{c \neq c_{i}}e_{c}$
and $\bp^{\top}v_{1} = (K-1)(a-b)$, so
\begin{align*}
  \bH(\bx_{i})v_{1}
  &= (K-1)a\,e_{c_{i}} - b\!\sum_{c \neq c_{i}}\!e_{c}
     \;-\; (K-1)(a-b)\bigl[a\,e_{c_{i}} + b\!\sum_{c \neq c_{i}}\!e_{c}\bigr] \\
  &= e_{c_{i}}\,(K-1)a\bigl[1 - (a-b)\bigr]
     \;+\; \sum_{c \neq c_{i}}e_{c}\,(-b)\bigl[1 + (K-1)(a-b)\bigr] \\
  &= e_{c_{i}}\,(K-1)a\,(K b)
     \;-\; \sum_{c \neq c_{i}}e_{c}\,b\,(K a)
   = K a b\,v_{1}.
\end{align*}
The two eigenvalues $Kab$ (on $V_{1}^{(i)}$) and $b$ (on
$V_{2}^{(i)}$) are equal iff $K a = 1$, i.e., $a = 1/K$ (uniform
prediction), which is excluded at non-trivial NC. The earlier draft
of this paper claimed these eigenvalues are equal by simplex-ETF
symmetry; that claim is incorrect because the stabilizer of class
$c_{i}$ is only $S_{K-1}$, not transitive on $\mathcal{Z}_{K}$.

\emph{Step 3: Eigenstructure of the per-sample $\widetilde\bGamma$
block.}
By Assumption~\ref{ass:nc-backbone}, the gradient features
$\widetilde\psi_{j,k}(\bx)$ commute with the $S_{K-1}$-action that
permutes the non-target indices $c \neq c_{i}$ (the simplex ETF and
the backbone are both invariant under this action; only the target
index $c_{i}$ is distinguished). Hence the per-sample block of
$\widetilde\bGamma$ on $\mathcal{Z}_{K}$ commutes with this
$S_{K-1}$-action and, by Schur's lemma applied to the
$S_{K-1}$-isotypic decomposition $\mathcal{Z}_{K} = V_{1}^{(i)} \oplus V_{2}^{(i)}$
(both isotypic components carry inequivalent
$S_{K-1}$-representations: $V_{1}^{(i)}$ is the trivial rep and
$V_{2}^{(i)}$ is the standard $(K-2)$-dimensional irreducible rep
of $S_{K-1}$), is block-diagonal with eigenvalues
$\gamma_{1}^{(i)} \geq 0$ on $V_{1}^{(i)}$ (multiplicity $1$) and
$\gamma_{2}^{(i)} \geq 0$ on $V_{2}^{(i)}$ (multiplicity $K-2$).
We assume $\gamma_{1}^{(i)}, \gamma_{2}^{(i)} > 0$ (i.e., the
gradient features genuinely populate both blocks); this is generic.

\emph{Step 4: Per-sample $r_{\mathrm{eff}}$ formula.}
The per-sample block $\bM_{\mathrm{CE}}^{(i)} = \widetilde\bSigma^{-1}\widetilde\bGamma\widetilde\bSigma^{-1}$
restricted to $\mathcal{Z}_{K}$ is, in the basis $V_{1}^{(i)} \oplus V_{2}^{(i)}$,
the diagonal matrix with entries $\gamma_{1}^{(i)}/(K a b)$ on
$V_{1}^{(i)}$ and $\gamma_{2}^{(i)}/b$ on $V_{2}^{(i)}$. Calling
these $\lambda_{1}^{(i)}$ and $\lambda_{2}^{(i)}$ respectively,
\[
  r_{\mathrm{eff}}(\bM_{\mathrm{CE}}^{(i)})
  = \frac{(\lambda_{1}^{(i)} + (K-2)\lambda_{2}^{(i)})^{2}}
         {(\lambda_{1}^{(i)})^{2} + (K-2)(\lambda_{2}^{(i)})^{2}}.
\]
Cauchy--Schwarz applied to the vectors $(1, \sqrt{K-2})$ and
$(\lambda_{1}^{(i)}, \sqrt{K-2}\,\lambda_{2}^{(i)})$ gives the
upper bound $1 + (K-2) = K-1$, attained iff
$\lambda_{1}^{(i)} = \lambda_{2}^{(i)}$. As $\lambda_{1}^{(i)}/\lambda_{2}^{(i)} \to \infty$
the ratio tends to $1$; as $\lambda_{1}^{(i)}/\lambda_{2}^{(i)} \to 0$ it tends
to $K-2$. Hence the lower bound is $1$ for $K \geq 3$. This proves~(a).

\emph{Step 5: Averaged operator is scalar by Schur's lemma.}
The class-averaged block is
\(
  \overline{\bM}_{\mathrm{CE}}
  = \frac{1}{n}\sum_{i=1}^{n}\bM_{\mathrm{CE}}^{(i)}
\)
with class-balanced training data ($n/K$ samples per class). By the
simplex-ETF symmetry of NC (Assumption~\ref{ass:nc-backbone}\,(a)--(b)),
the data distribution and the predictor are invariant under the
full permutation group $S_{K}$ acting simultaneously on classes,
last-layer features, and rows of $W$. The per-sample blocks
$\bM_{\mathrm{CE}}^{(i)}$ depend on $i$ only through the class
$c_{i}$, and the averaged operator $\overline{\bM}_{\mathrm{CE}}$
commutes with the diagonal $S_{K}$-action on $\R^{K}$.

The standard fact about the $(K-1)$-dimensional zero-sum subspace
$\mathcal{Z}_{K} \subset \R^{K}$ is that it is the
\emph{standard irreducible representation} of $S_{K}$ (see, e.g.,
\cite[\S 2.3]{Amari2016} or any text on symmetric-group
representations). By Schur's lemma, every $S_{K}$-equivariant
endomorphism of an irreducible representation is a scalar. Hence
$\overline{\bM}_{\mathrm{CE}}|_{\mathcal{Z}_{K}} = \kappa\,\bP_{\mathcal{Z}_{K}}$
for some scalar $\kappa > 0$ (positivity follows from the fact that
each $\bM_{\mathrm{CE}}^{(i)}$ is positive semidefinite and at
least one per-sample block has $\gamma_{j}^{(i)} > 0$). Therefore
all $K-1$ nonzero eigenvalues of $\overline{\bM}_{\mathrm{CE}}$ are
equal, and by Definition~\ref{def:reff},
\[
  r_{\mathrm{eff}}(\overline{\bM}_{\mathrm{CE}})
  = \frac{((K-1)\kappa)^{2}}{(K-1)\kappa^{2}}
  = K-1.
\]
This proves~(b).
\end{proof}

\begin{remark}[How does class-averaging restore $K-1$?]\label{rem:averaging}
The mechanism is purely group-theoretic. Each per-sample block
$\bM_{\mathrm{CE}}^{(i)}$ has only an $S_{K-1}$ symmetry (the
stabilizer of class $c_{i}$), and $\mathcal{Z}_{K}$ decomposes as
$V_{1}^{(i)} \oplus V_{2}^{(i)}$ with two distinct eigenvalues.
But the target direction $V_{1}^{(c)}$ moves as $c$ varies over the
$K$ classes: explicitly, $V_{1}^{(c)} = \spn(K e_{c} - \mathbf{1})$
and one computes
\[
  \sum_{c=1}^{K}
  \frac{(K e_{c} - \mathbf{1})(K e_{c} - \mathbf{1})^{\top}}{\norm{Ke_{c} - \mathbf{1}}^{2}}
  \Bigg|_{\mathcal{Z}_{K}}
  \;=\; \frac{K}{K-1}\,\bP_{\mathcal{Z}_{K}}.
\]
That is, summing the rank-one projectors onto target directions
across all classes yields a scalar on $\mathcal{Z}_{K}$. Combined
with the corresponding sum over the $V_{2}^{(c)}$-blocks (also
scalar by symmetry), the class-average produces an operator with a
single eigenvalue across all $K-1$ directions. Schur's lemma is the
clean abstract way to phrase this.
\end{remark}

\begin{remark}[Dimension-free scaling for
classification]\label{rem:dim-free}
Theorem~\ref{thm:ce-rank}\,(b) states that, at the NC fixed point,
the effective rank of the \emph{class-averaged} block
$\overline{\bM}_{\mathrm{CE}}$ depends only on $K$ and not on the
input dimension $d$ or the parameter count $p$. This is the
classification analog of the dimension-free regression rate of
\cite{paperC} (rank-one collapse onto a single target direction):
both rest on a residual symmetry of the trained predictor at its
fixed point that quotients out the ambient parameter dimension. In
classification the residual symmetry is $S_{K}$ acting on the
simplex-ETF directions, and the target subspace
$\mathcal{Z}_{K} = \R^{K}/\spn(\mathbf{1})$ is irreducible of
dimension $K-1$.
\end{remark}

\begin{remark}[Full-operator effective rank]\label{rem:full-vs-per-sample}
With $n$ training samples class-balanced and the $S_{K}$-symmetry
of NC, the full operator $\bM_{\mathrm{CE}}$ on the joint
sample-class space has rank $n(K-1)$ on the zero-sum subspace and
$r_{\mathrm{eff}}(\bM_{\mathrm{CE}}) = n(K-1)$ at the limit point
(once both the per-sample two-block structure and the $S_{K}$-
averaging across classes are simultaneously taken into account).
The dimension-free scaling we care about for generalization is the
class-averaged quantity, which captures the noise leakage \emph{per
training example averaged over class identity}; this is the right
object because the noise-leakage variance
$\sigma^{2}\tr(\bM_{\mathrm{CE}})$ from
Lemma~\ref{lem:noise-leakage} aggregates over both samples and
classes and is invariant under the diagonal $S_{K}$-action.
\end{remark}

\subsection{Comparison with prior work}\label{subsec:prior-work}

The result of Theorem~\ref{thm:ce-rank} is positioned at the
intersection of three prior literatures, and its novelty lies
precisely in the operator-framework angle that none of them
addresses.
\begin{itemize}
\item \emph{NC geometry (Lu--Steinerberger~\cite{lu2022},
Mixon--Parshall--Pi~\cite{mixon2022}).} These works characterize
the NC fixed point itself in the unconstrained-features model
under cross-entropy: the simplex ETF, self-duality, and within-class
collapse. They establish that NC is the unique global minimizer of
the layer-peeled / unconstrained-features objective, and they
quantify how rapidly NC is approached. We use their geometric
characterization as input (Assumption~\ref{ass:nc-backbone}) and
do not re-derive it.
\item \emph{Central path under MSE (Han--Papyan--Donoho~\cite{han2022}).}
This work studies the dynamical approach to NC under squared loss,
identifying a central path in feature/classifier space. Our
treatment is for the cross-entropy noise-propagation operator
$\bM_{\mathrm{CE}}$, not for the central-path dynamics; the
loss-specific ingredient (Fisher matrix $\bH$ instead of identity)
is what forces the two-eigenvalue per-sample structure of
Theorem~\ref{thm:ce-rank}\,(a) and the $S_{K}$-equivariant Schur
argument of Theorem~\ref{thm:ce-rank}\,(b).
\item \emph{Class-imbalanced extension (Fang et al.~\cite{Fang2021}).}
The unconstrained-features model under class imbalance shows that
the simplex ETF deforms to a frequency-weighted variant. Our
$S_{K}$-symmetry argument relies on class balance; the imbalanced
case would replace $S_{K}$-equivariance by a smaller invariance
group and is left to future work.
\end{itemize}
The novel contribution here is the
\textbf{$r_{\mathrm{eff}}$-of-$\bM_{\mathrm{CE}}$ characterization
within the noise-propagation operator framework of~\cite{paperA}}:
identifying the right object (the class-averaged block on
$\mathcal{Z}_{K}$, not the per-sample block), exhibiting the exact
two-eigenvalue per-sample formula, and using $S_{K}$-irreducibility
of $\mathcal{Z}_{K}$ to obtain $r_{\mathrm{eff}} = K-1$ for the
averaged operator. The earlier draft of this paper conflated the
per-sample and averaged statements; the corrected statement
delineates them, and is what is actually proven by simplex-ETF
geometry plus Schur's lemma.

%% ============================================================
\section{Discussion}\label{sec:disc}
%% ============================================================

\paragraph{Status of the NC hypothesis.}
The conclusion $r_{\mathrm{eff}}(\overline{\bM}_{\mathrm{CE}}) = K-1$
(class-averaged block) is conditional on the NC fixed point being
attained or approached at the end of training. NC is observed
empirically across diverse networks~\cite{papyan2020}, and is
rigorously established in the unconstrained-features model
of~\cite{lu2022,mixon2022}, which idealizes the last-layer features
as free variables. The result of this paper is therefore a
statement of the form ``NC $+$ backbone non-degeneracy
(Assumption~\ref{ass:nc-backbone}(c)) $\Rightarrow$ rank-$(K-1)$
effective rank for the class-averaged block''; we do not prove that
NC holds for finite networks trained end-to-end with cross-entropy.
Per-sample blocks have $r_{\mathrm{eff}}^{(i)} \in [1, K-1]$ in
general (see Theorem~\ref{thm:ce-rank}\,(a) and Step~2 of the
proof); the dimension-free $K-1$ scaling is a property of the
class-averaged object only.
Limitations in the strict one-hot limit are addressed by rescaling
(Remark~\ref{rem:strict-NC}); the strict limit does \emph{not}
flatten the per-sample two-block spectrum, but the class-averaged
quantity is unaffected.

\paragraph{Off-fixed-point behavior.}
A natural follow-up question is: how does
$r_{\mathrm{eff}}(\bM_{\mathrm{CE}})$ behave during the approach to
the NC fixed point? We expect a gap term whose decay is governed by
the convergence rate of NC, which would yield a rate of
$r_{\mathrm{eff}}(\bM_{\mathrm{CE}})(t) - (K-1)$ in terms of the
training time $t$. Quantifying this gap is open.

\paragraph{Class-imbalanced data.}
Theorem~\ref{thm:ce-rank} assumes class-balanced gradient features.
For imbalanced classes, the simplex ETF deforms to a
class-frequency-weighted variant (see e.g.~\cite{Fang2021} for the
unconstrained-features analysis), and the effective rank gains a
class-frequency-dependent correction. We do not pursue this here.

\paragraph{Connection to the regression result of~\cite{paperC}.}
The companion paper~\cite{paperC} (this paper's regression
counterpart) shows that, after feature-direction alignment, the
effective rank of the regression operator
$\bM = \bSigma^{-1}\bGamma\bSigma^{-1}$ collapses to a single
dimension along the target direction. In classification, the
analogous collapse for the class-averaged block
$\overline{\bM}_{\mathrm{CE}}$ is to $K-1$ dimensions along the
simplex faces: each class identifies its corresponding simplex
vertex, and the rank-one regression collapse becomes a
rank-$(K-1)$ classification collapse. Both reflect the same
abstract mechanism --- a residual symmetry of the trained
predictor at its fixed point reduces an a priori $\Theta(d)$ rank
to a small representation-theoretic invariant --- but the per-sample
behavior differs: paper~\cite{paperC}'s per-sample rank-one
collapse is exact for each sample, whereas in classification the
per-sample blocks have only the smaller stabilizer $S_{K-1}$ and
the dimension-free $K-1$ statement holds only after $S_{K}$-averaging
across class identities. We retain distinct notation
$\bM$ vs.\ $\bM_{\mathrm{CE}}$ throughout to disambiguate the two
operators.

\subsection*{Acknowledgements}

The exploratory analysis underlying Theorems~\ref{thm:ce-trichotomy}
and~\ref{thm:ce-rank}, including the derivation of the
Fisher-corrected operator from the cross-entropy Hessian, was
conducted with the assistance of Claude (Anthropic). All
mathematical statements and proofs have been verified by the
author.

%% ============================================================
%% References
%% ============================================================

\begin{thebibliography}{20}

\bibitem{paperA}
L.~Chang.
\newblock Noise propagation operators and a two-parameter characterization
  of benign overfitting.
\newblock \emph{Companion paper}, 2026.

\bibitem{paperB}
L.~Chang.
\newblock A bias--variance decomposition for the noise propagation operator
  and the role of feature learning.
\newblock \emph{Companion paper}, 2026.

\bibitem{paperC}
L.~Chang.
\newblock Feature learning, effective dimension, and architecture-specific
  critical depth.
\newblock \emph{Companion paper}, 2026.

\bibitem{Amari2016}
S.-i.~Amari.
\newblock \emph{Information Geometry and Its Applications}.
\newblock Springer, 2016.

\bibitem{Fang2021}
C.~Fang, H.~He, Q.~Long, and W.~J.~Su.
\newblock Exploring deep neural networks via layer-peeled model:
  Minority collapse in imbalanced training.
\newblock \emph{Proceedings of the National Academy of Sciences},
  118(43):e2103091118, 2021.

\bibitem{han2022}
X.~Y.~Han, V.~Papyan, and D.~L.~Donoho.
\newblock Neural collapse under MSE loss: proximity to and dynamics on
  the central path.
\newblock In \emph{International Conference on Learning Representations
  (ICLR)}, 2022.

\bibitem{lu2022}
J.~Lu and S.~Steinerberger.
\newblock Neural collapse under cross-entropy loss.
\newblock \emph{Applied and Computational Harmonic Analysis},
  59:224--241, 2022.

\bibitem{mei2022}
S.~Mei, T.~Misiakiewicz, and A.~Montanari.
\newblock Generalization error of random feature and kernel methods:
  Hypercontractivity and kernel matrix concentration.
\newblock \emph{Applied and Computational Harmonic Analysis},
  59:3--84, 2022.

\bibitem{mixon2022}
D.~G.~Mixon, H.~Parshall, and J.~Pi.
\newblock Neural collapse with unconstrained features.
\newblock \emph{Sampling Theory, Signal Processing, and Data Analysis},
  20(2):11, 2022.

\bibitem{papyan2020}
V.~Papyan, X.~Y.~Han, and D.~L.~Donoho.
\newblock Prevalence of neural collapse during the terminal phase of deep
  learning training.
\newblock \emph{Proceedings of the National Academy of Sciences},
  117(40):24652--24663, 2020.

\end{thebibliography}

\end{document}
